From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-08 Pi+/paper/discussion.tex
Fixes in sec finite
working on sec 8

\section{Discussion \& Related Work}~\label{sec:discussion}

In this paper, we \ldots

@@

% Let's think of the $\PiLang$ combinators as describing the inhabitants of the dentity type of finite types.

By giving a computable presentation for a univalent subuniverse, we are able to describe the path space of it
syntactically, by giving a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom. 
The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
The $\term{ua}$ map corresponds to synthesing a program from an equivalence (which, in general, is of course
undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
a constructive argument. Then, the computation rule $\term{ua-\beta}$ expresses the fact that program synthesis is
sound, while $\term{ua-\eta}$ corresponds to the soundess of the equational theory ($\PiLang$ 2-combinators). Thus, we
this suggests a new computational interpretation of the univalence principle, which provides an intution on why certain
constructions are hard (or impossible in general case).

@@

\begin{center}
    \begin{tabular}{c c c}
        \hline
        $\PiLang$     & $\Sn$               & $\UFin$     \\
        \hline
        types         & Natural numbers     & Finite sets \\
        \hline
        1-combinators & Generators of $\Sn$ & 1-paths     \\
        \hline
        2-combinators & Relations of $\Sn$  & 2-paths     \\
        \hline
    \end{tabular}
\end{center}

\begin{center}
    \begin{tabular}{c c c}
        \hline
        Logic                              & Program                       & Algebra                     \\
        \hline
        Propositions                       & Types                         & Objects                     \\
        \hline
        Proofs                             & Terms                         & Morphisms                   \\
        \hline
        $\blank$                           & Equations                     & Homotopies                  \\
        \hline
        Intuitionistic propositional logic & Simply typed Lambda Calculus  & Cartesian Closed Categories \\
        \hline
        Superstructural Reversible logic   & Reversible Language $\PiLang$ & Symmetric Rig Groupoids     \\
        \hline
        Connectives                        & Type constructors             & Bifunctors                  \\
        \hline
        ?                                  & Type normalization            & ?                           \\
        \hline
        ?                                  & Program normalization         & ?                           \\
        \hline
        ?                                  & Soundness of 2-combinators    & ?                           \\
    \end{tabular}
\end{center}

\begin{center}
    \begin{tabular}{c c c}
        \hline
        Reversible Logic             & Reversible Programming                  & Algebra                  \\
        \hline
        Free commutative rig         & Free symmetric rig groupoid             & $\UFin$                  \\
        \hline
        Propositions                 & Types                                   & Objects                  \\
        \hline
        Entailment?                  & Reversible programs                     & Isomorphisms             \\
        \hline
        $\blank$                     & 2-combinators                           & Homotopies               \\
        \hline
        Conjunction?                 & $\times$                                & $\otimes$                \\
        \hline
        Disjunction?                 & $+$                                     & $\oplus$                 \\
        \hline
        $A \land B \vdash B \land A$ & $\swapt$                                & $\mathcal{B}_{A,B}$      \\
        \hline
        True proposition             & $\onet$                                 & $F_1$                    \\
        \hline
        Decidability                 & Program Synthesis                       & Reachability in a graph? \\
        \hline
        Soundness/Completeness       & Soundness/Completeness of 1-combinators & ?                        \\
    \end{tabular}
\end{center}

@@

>>>>

\section{Discussion \& Related Work}
\label{sec:discussion}

% In this paper, we \ldots

@@

% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.

By giving a computable presentation for a univalent subuniverse, we are able to describe the path space of it
syntactically, by giving a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which, in general, is of course
undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
a constructive argument. Then, the computation rule $\term{ua-\beta}$ expresses the fact that program synthesis is
sound, while $\term{ua-\eta}$ corresponds to the soundness of the equational theory ($\PiLang$ 2-combinators). Thus, we
this suggests a new computational interpretation of the univalence principle, which provides an intuition on why certain
constructions are hard (or impossible in general case).

@@

% \begin{center}
%     \begin{tabular}{c c c}
%         \hline
%         $\PiLang$     & $\Sn$               & $\UFin$     \\
%         \hline
%         types         & Natural numbers     & Finite sets \\
%         \hline
%         1-combinators & Generators of $\Sn$ & 1-paths     \\
%         \hline
%         2-combinators & Relations of $\Sn$  & 2-paths     \\
%         \hline
%     \end{tabular}
% \end{center}

% \begin{center}
%     \begin{tabular}{c c c}
%         \hline
%         Logic                              & Program                       & Algebra                     \\
%         \hline
%         Propositions                       & Types                         & Objects                     \\
%         \hline
%         Proofs                             & Terms                         & Morphisms                   \\
%         \hline
%         $\blank$                           & Equations                     & Homotopies                  \\
%         \hline
%         Intuitionistic propositional logic & Simply typed Lambda Calculus  & Cartesian Closed Categories \\
%         \hline
%         Superstructural Reversible logic   & Reversible Language $\PiLang$ & Symmetric Rig Groupoids     \\
%         \hline
%         Connectives                        & Type constructors             & Bifunctors                  \\
%         \hline
%         ?                                  & Type normalisation            & ?                           \\
%         \hline
%         ?                                  & Program normalisation         & ?                           \\
%         \hline
%         ?                                  & Soundness of 2-combinators    & ?                           \\
%     \end{tabular}
% \end{center}

% \begin{center}
%     \begin{tabular}{c c c}
%         \hline
%         Reversible Logic             & Reversible Programming                  & Algebra                  \\
%         \hline
%         Free commutative rig         & Free symmetric rig groupoid             & $\UFin$                  \\
%         \hline
%         Propositions                 & Types                                   & Objects                  \\
%         \hline
%         Entailment?                  & Reversible programs                     & Isomorphisms             \\
%         \hline
%         $\blank$                     & 2-combinators                           & Homotopies               \\
%         \hline
%         Conjunction?                 & $\times$                                & $\otimes$                \\
%         \hline
%         Disjunction?                 & $+$                                     & $\oplus$                 \\
%         \hline
%         $A \land B \vdash B \land A$ & $\swapt$                                & $\mathcal{B}_{A,B}$      \\
%         \hline
%         True proposition             & $\onet$                                 & $F_1$                    \\
%         \hline
%         Decidability                 & Program Synthesis                       & Reachability in a graph? \\
%         \hline
%         Soundness/Completeness       & Soundness/Completeness of 1-combinators & ?                        \\
%     \end{tabular}
% \end{center}

@@

\todo{Why is this better than evaluating the permutation and checking each bit?}
\todo{Why care about an equational theory? Compilers, optimisation.}
